---
output: 
  github_document:
    number_sections: false
editor_options: 
  chunk_output_type: console
---

<!-- README.md is generated from README.Rmd. Please edit README.Rmd and run devtools::build_readme() -->
# EloRating.Bayes

## Installation

You need [`cmdstanr`](https://mc-stan.org/cmdstanr/) in order to install and run `EloRating.Bayes`. 
This in turn requires a working C++ toolchain first. 
Check out the [getting started-guide](https://mc-stan.org/cmdstanr/articles/cmdstanr.html) from `cmdstanr` to see how to get everything set up properly. 
Also [this document](https://mc-stan.org/docs/cmdstan-guide/cmdstan-installation.html#cpp-toolchain) might be helpful.

Then install `cmdstanr`.

```{r, eval = FALSE}
install.packages("cmdstanr", repos = c("https://mc-stan.org/r-packages/", getOption("repos")))
```

And you also need the `remotes` packages, which is easy to install from CRAN:

```{r, eval = FALSE}
install.packages("remotes")
```


Then check whether things are set up correctly:

```{r, eval = FALSE}
library(cmdstanr)
check_cmdstan_toolchain(fix = TRUE)
```

If this gives positive feedback, install `EloRating.Bayes`:

```{r, eval = FALSE}
library(remotes)
remotes::install_github("gobbios/EloRating.Bayes", dependencies = TRUE, build_vignettes = FALSE)
```


If you want to install (recompile) the intro vignette, use:

```{r, eval = FALSE}
library(remotes)
remotes::install_github("gobbios/EloRating.Bayes", dependencies = TRUE, build_vignettes = TRUE)
```


## Examples

```{r, include=FALSE}
library(EloRating.Bayes)
```

```{r, eval=FALSE}
res <- toydata()
pp_check(res)
```

```{r pp_plot_ex01, echo=FALSE}
res <- toydata()
par(family = "serif", mgp = c(1.5, 0.4, 0), tcl = 0.2, mar = c(2.5, 2.5, 1.5, 0.5))
set.seed(1)
pp_check(res)
```



