---
output: 
  github_document:
    number_sections: false
editor_options: 
  chunk_output_type: console
---

<!-- README.md is generated from README.Rmd. Please edit README.Rmd and run devtools::build_readme() -->
# EloRating.Bayes

## Installation

You need [`cmdstanr`](https://mc-stan.org/cmdstanr/) in order to install and run `EloRating.Bayes`. 
This in turn requires a working C++ toolchain first. 
Check out the [getting started-guide](https://mc-stan.org/cmdstanr/articles/cmdstanr.html) from `cmdstanr` to see how to get everything set up properly. 
Also [this document](https://mc-stan.org/docs/cmdstan-guide/cmdstan-installation.html#cpp-toolchain) might be helpful.

Then install `cmdstanr`.

```{r, eval = FALSE}
install.packages("cmdstanr", repos = c("https://mc-stan.org/r-packages/", getOption("repos")))
```

And you also need the `remotes` packages, which is easy to install from CRAN:

```{r, eval = FALSE}
install.packages("remotes")
```


Then check whether things are set up correctly:

```{r, eval = FALSE}
library(cmdstanr)
check_cmdstan_toolchain(fix = TRUE)
```

If this gives positive feedback, install `EloRating.Bayes`:

```{r, eval = FALSE}
library(remotes)
remotes::install_github("gobbios/EloRating.Bayes", dependencies = TRUE, build_vignettes = FALSE)
```


If you want to install (recompile) the intro vignette, use:

```{r, eval = FALSE}
library(remotes)
remotes::install_github("gobbios/EloRating.Bayes", dependencies = TRUE, build_vignettes = TRUE)
```


## Examples

```{r, include=FALSE}
library(EloRating.Bayes)
```


```{r, eval=FALSE}
# create a toy data set
set.seed(123)
x <- EloRating::randomsequence(nID = 6, avgIA = 10, presence = c(0.4, 0.4))
winner <- x$seqdat$winner
loser <- x$seqdat$loser
Date <- x$seqdat$Date
presence <- x$pres
intensity <- sample(c("mild", "severe"), length(winner), TRUE)

# prep data
standat <- prep_seq(winner = winner, loser = loser, Date = Date, presence = presence, intensity = intensity)

# fit 
res <- elo_seq_bayes(standat = standat, quiet = FALSE, parallel_chains = 4, seed = 1)
```

A numeric summary.

```{r, eval=TRUE, echo=2:2}
res <- toydata()
summary(res)
```

Example of posterior predictive check: proportion of won interactions per individual.

```{r, eval=FALSE}
pp_check(res, n_samples = 50)
```

```{r pp_plot_ex01, echo=FALSE, fig.width=7, fig.height=2.7, out.width="70%"}
par(family = "serif", mgp = c(1.5, 0.4, 0), tcl = 0.2, mar = c(2.5, 2.5, 1.5, 0.5))
set.seed(1)
pp_check(res, n_samples = 50)
```

Plot posteriors of ratings. By default this is on the last day of data set (here one individual is absent that day, *o*, and hence missing from the plot).

```{r, eval=FALSE}
plot_scores(res)
```

```{r scores_01, echo=FALSE, fig.width=7, fig.height=2.7, out.width="70%"}
par(family = "serif", mgp = c(1.2, 0.4, 0), tcl = 0.2, mar = c(2.9, 2.5, 1.5, 0.5))
set.seed(1)
plot_scores(res)
```


